---
title: "compbio_proj"
author: "David Lim, Mark He, Mike Lafferty"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---
#Introduction:
<Mike>


#Figures:
<David>

Figure 1C:

```{r}
library(readxl)
library(genefilter)

X <- read_excel("rnaseq_CNS_all.xlsx")

# filter out genes with 0 SD (or 0 variance)
X.SDs <- rowSds(X[,-1])
good = (X.SDs!=0)

X = X[good,]

cor_mat = cor(X[,-1],method="spearman")
library(pheatmap)
pheatmap(cor_mat,cluster_rows=F,cluster_cols=F,main="Spearman Correlation of Cell Lines")
```

- High correlation within cell line replicates (reproducibility of RNA-seq)
- Normalization of correlation (?!)
- Whole brain: correlated with Neuron & OPC
- NFO & MO & OPC correlated (same lineage)

Figure 1D:

```{r}
sel_genes = c("Gfap","Aldh1l1","Slc1a3","Aqp4","Tubb3",
              "Stmn2","Snap25","Eno2","Syn1","Pdgfra",
              "Cspg4","Enpp6","Nfasc","Pip","Mog","Mbp",
              "Mobp","Ccl3","Itgam","Tnf","Cldn5","Flt1","Esam")    # Itgam = Cd11b in paper

sel_X <- X[X$gene.symbol %in% sel_genes,]  # subsetting genes and omitting WB1,2,3
sel_X <- sel_X[c(10,1,17,2,22,19,18,6,20,16,5,7,15,14,12,13,3,11,21,4,9,8),]
sel_Xcts <- sel_X[,-1]                       # subset just the counts

sel_means <- rowMeans(sel_Xcts)
sel_sd <- rowSds(sel_Xcts)


num_sd <- matrix(0,nrow=nrow(sel_Xcts),ncol=ncol(sel_Xcts))
for(i in 1:nrow(sel_Xcts)){
  num_sd[i,] <- (as.numeric(sel_Xcts[i,]) - sel_means[i])/sel_sd[i]    # calculate # of SD's away from rowMean
}

rownames(num_sd) <- as.matrix(sel_X[,1])
colnames(num_sd) <- colnames(sel_X[,-1])

pheatmap(num_sd,cluster_cols=F,cluster_rows=F,main="Number of SDs Away From Mean Across Cell Lines")

```

- Gene "Cd11b" not found. Different name: "Itgam"
- Gene "Pip"" yielded SD = 0 --> omitted.
- High SD of astrocytes and neurons coincide with higher SD of whole brain



Figure 3B:

```{r}
all_genes <- X[,1]
Xcts <- X[,2:15]
enrich <- matrix(0,nrow=nrow(Xcts),ncol=ncol(Xcts))             # store enrichment values

for(i in 1:nrow(Xcts)){
  enrich[i,]<-as.numeric(Xcts[i,])/sum(as.numeric(Xcts[i,]))
}

enrich2 <- matrix(0,nrow=2*nrow(enrich),ncol=ncol(enrich)/2)         # new matrix to merge replicates
for(j in 1:(ncol(enrich)/2)){
  enrich2[1:nrow(enrich),j] <- enrich[,(2*j-1)]
  enrich2[(nrow(enrich)+1):(2*nrow(enrich)),j] <- enrich[,(2*j)]
}

enrich_genes <- matrix(0,nrow=40,ncol=ncol(Xcts))
enrich_id <- NULL

for(j in 1:ncol(enrich2)){
  #enrich_genes[,j] <- as.matrix(all_genes[order(-enrich[,j])[1:40],1])
  enrich_id <- c(enrich_id,order(-enrich2[,j])[1:40])
}

for(i in 1:length(enrich_id)){
  if(enrich_id[i] > nrow(X)){
    enrich_id[i] = enrich_id[i] - nrow(X)
  }
}

enrich_X <- X[enrich_id,2:15]

enrich_means <- rowMeans(enrich_X)
enrich_sd <- rowSds(enrich_X)

num_sd <- matrix(0,nrow=nrow(enrich_X),ncol=ncol(enrich_X))

for(i in 1:nrow(enrich_X)){
  num_sd[i,] <- (as.numeric(enrich_X[i,]) - enrich_means[i])/enrich_sd[i]    # calculate # of SD's away from rowMean
}

colnames(num_sd) = colnames(enrich_X)


dists = dist(num_sd)
hc = hclust(dists)

o = hc$order[c(41:120,1:40,241:280,201:240,161:200,121:160)]
Y = t(num_sd)[,o]


pheatmap(Y,cluster_rows=F,cluster_cols=F,show_colnames=F,main="Number of SDs Away for Top 40 Enriched Genes Per Cell Line")

```

- Top 40 "enriched" genes for each cell line
- Is "enrichment" a good indicator of genes that discrimate across cell lines?
- Selecting top 40 for every cell line: genes with smaller variation may be included
- Maybe interesting: top 300 variance genes

```{r}
# Using genes with top 100 row variances
std.dev <- rowSds(X[,2:15])
sel.X <- X[order(-std.dev)[1:300],]

genes <- sel.X[,1]
sel.Xcts <- sel.X[,2:15]

sel_means <- rowMeans(sel.Xcts)
sel_sd <- rowSds(sel.Xcts)

num_sd <- matrix(0,nrow=nrow(sel.Xcts),ncol=ncol(sel.Xcts))

for(i in 1:nrow(sel.Xcts)){
  num_sd[i,] <- (as.numeric(sel.Xcts[i,]) - sel_means[i])/sel_sd[i]    # calculate # of SD's away from rowMean
}

rownames(num_sd) = as.matrix(genes)
colnames(num_sd) = colnames(sel.Xcts)


dists = dist(num_sd)
hc = hclust(dists)

Y = t(num_sd)[,hc$order]


pheatmap(Y,cluster_rows=F,cluster_cols=F,show_colnames=F,main="Number of SDs Away for Top 300 Variance Genes")


```


- Variation paired by replicates (as expected)
- Most highly variable genes seem to be specific to microglial cell lines
- Correlation between genes specific to NFO & MO (newly-formed vs. myelinating oligodendrocytes)





#Correlating 2 cell types:
<Mark>


#Cibersort:
<Mike>


#Conclusions:



